Estimating Learning Dynamics in the Routing Game

### 2015-10-14 paper/paper.tex
Updated paper.

%-----------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Estimating the decay rate of the learning rate sequence}
In the previous section, we proposed a method to estimate one term of the learning rate sequence. One can of course repeat this procedure at every iteration, thus generating a sequence of estimated learning rates. However, the resulting sequence may not be decreasing. In order to be consistent with the assumptions of the model, we can assume a parameterized sequence of learning rates (which is by construction decreasing), then estimate the parameters of the sequence, given the observations. Motivated by Theorem~\ref{thm:convergence}, we will assume, in this section, that $\eta^{(t)}_k = \eta^{(0)}_k t^{-\alpha_k}$ with $\eta^{(0)}_k > 0$ and $\alpha_k \in (0, 1)$.

@@

%-----------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Traffic flow prediction}
We discuss one important application of the proposed estimation problem. Once we have estimated the learning rates, we can propagate the model forward in order to predict the distributions of the players for the next time step. More precisely, if at iteration $t$, we have observed $\bar x^{(t)}$, $\bar \ell^{(t)}$, and we have estimated the terms $(\eta_k^{(1)}, \dots, \eta_k^{(t-1)})$ for a player $k$, then we can use these terms to estimate $\eta_k^{(t)}$, and predict the next distribution by solving
\al{
x_k^{(t+1)}
&= \argmin{x_k \in \Delta^{\Pcal_k}} \braket{\eta_k^{(t)} }{ \ell(\bar x^{(t)}_k) } + D_{\psi_k}(x_k, \bar x^{(t)}_k) \\
&\overset{\Delta}{=} g(\bar x^{(t)}_k, \eta_k^{(t)}),
}
where we defined the function $g$, which takes a distribution and a learning rate and propagates the model forward one step. We can inductively estimate the next terms by propagating the model further:
\[
x_k^{(t+i+1)} = g(x^{(t+i)}_k, \eta_k^{(t+i)}).
\]
Here, we assume that we can extrapolate the learning rate sequence to estimate the terms $\eta_k^{(t+i)}$. If we assume a particular form of the sequence, $\eta_k^{(t)} = \eta_k^{(0)} t^{-\alpha_k}$, then this can be done readily once we have an estimate of $\eta_k^{(0)}$ and $\alpha_k$. However, if each term of the sequence is estimated separately, we need to use model to predict the next terms. We propose a few simple methods in Section~\ref{sec:results}, and evaluate their generalization performance.

@@

@@

\section{Analysis}
\label{sec:analysis}

@@

At the beginning of the game, there is a clear exploration phase where players tend make aggressive assignments with their flow distribution. Towards the end of the game, the players consolidate their position and tend to make more conservative flow distribution assignments. An example of this behavior can be seen in Figure~\ref{fig:player_behavior}.

We can describe the convergence of the system more concretely by evaluating the Rosenthal potential of the game at each turn from Prop~\ref{prop:potential}. From Figure~\ref{fig:global_potential}, the potential of the system oscillates rapdily in the exploration phase and then converges to the minimal potential. The system also shows robustness seen in Figure~\ref{fig:global_potential} where it recovers from perturbations in the convergence phase.

@@

We can estimate the learning rate of each player with the update in Eq~\ref{eq:estimation_eta}. We obtained these sequence of learning rates for each player and for better visualization, we plot the moving average

% \begin{figure}
%   \centering
%   \includegraphics[width=0.5\textwidth]{images/learning_rates}
%   \caption{Learning rate for each player.}
%   \label{fig:learning_rates}
% \end{figure}

With these estimated sequence of learning rates for each player, we can predict the flow distribution at turn $t+1$ with the learning rate $\eta_t$.
% \[
% \eta_t=\dfrac{1}{N}\sum_{i=t-N-1}^{t-1}\eta_i
% \] and finally we compute $\eta_t$ as a linear regression from $\eta_i,\ i \in [0,\ t-1]$.

We also compare these methods with a parameterized sequence of learning rate in Section 3.2

% \begin{figure}
%   \centering
%   \includegraphics[width=0.5\textwidth]{images/previous_eta_predictions}
%   \caption{Flow distribution predictions $\eta_t = \eta_{t-1}$.}
%   \label{fig:previous_eta_predictions}
% \end{figure}

% \begin{figure}
%   \centering
%   % \includegraphics[width=0.5\textwidth]{images/fill_me}
%   \missingfigure[figwidth=6cm]{Testing a long text string}
%   \caption{Flow distribution predictions moving average $\eta_t$.}
%   \label{fig:moving_eta_predictions}
% \end{figure}

% \begin{figure}
%   \centering
%   % \includegraphics[width=0.5\textwidth]{images/fill_me}
%   \missingfigure[figwidth=6cm]{Testing a long text string}
%   \caption{Flow distribution predictions linear regression $\eta_t$.}
%   \label{fig:linear_regression_eta_predictions}
% \end{figure}

% \begin{figure}
%   \centering
%   % \includegraphics[width=0.5\textwidth]{images/fill_me}
%   \missingfigure[figwidth=6cm]{Testing a long text string}
%   \caption{Flow distribution predictions parameterized $\eta_t$.}
%   \label{fig:parameterized_eta_predictions}
% \end{figure}

% \begin{figure}
%   \centering
%   \includegraphics[width=0.5\textwidth]{images/no_slider_actual_flow_distribution.png}
%   \caption{Player flow distribution.}
%   \label{fig:no_slider_player_flow_distribution}
% \end{figure}

% \begin{figure}
%   \centering
%   \includegraphics[width=0.5\textwidth]{images/no_slider_KL_divergence_surface}
%   \caption{KL Divergence surface.}
%   \label{fig:kl_divergence_surface}
% \end{figure}

>>>>

%-----------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Estimating the decay rate of the learning rate sequence}
\label{sec:estimation}
In the previous section, we proposed a method to estimate one term of the learning rate sequence. One can of course repeat this procedure at every iteration, thus generating a sequence of estimated learning rates. However, the resulting sequence may not be decreasing. In order to be consistent with the assumptions of the model, we can assume a parameterized sequence of learning rates (which is by construction decreasing), then estimate the parameters of the sequence, given the observations. Motivated by Theorem~\ref{thm:convergence}, we will assume, in this section, that $\eta^{(t)}_k = \eta^{(0)}_k t^{-\alpha_k}$ with $\eta^{(0)}_k > 0$ and $\alpha_k \in (0, 1)$.

@@

%-----------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Traffic flow prediction}
We discuss one important application of the proposed estimation problem. Once we have estimated the learning rates, we can propagate the model forward in order to predict the distributions of the players for the next time step. More precisely, if at iteration $t$, we have observed $\bar x^{(t)}$, $\bar \ell^{(t)}$, and we have estimated the terms $(\eta_k^{(1)}, \dots, \eta_k^{(t-1)})$ for a player $k$, then we can use these terms to estimate $\eta_k^{(t)}$, and predict the next distribution by solving

\begin{equation}
  \label{eq:estimation_step}
  \begin{aligned}
x_k^{(t+1)}
&= \argmin{x_k \in \Delta^{\Pcal_k}} \braket{\eta_k^{(t)} }{ \ell(\bar x^{(t)}_k) } + D_{\psi_k}(x_k, \bar x^{(t)}_k) \\
&\overset{\Delta}{=} g(\bar x^{(t)}_k, \eta_k^{(t)}),
\end{aligned}
\end{equation}

where we defined the function $g$, which takes a distribution and a learning rate and propagates the model forward one step. We can inductively estimate the next terms by propagating the model further:
\begin{equation}
\label{eq:propagation}
x_k^{(t+i+1)} = g(x^{(t+i)}_k, \eta_k^{(t+i)}).
\end{equation}
Here, we assume that we can extrapolate the learning rate sequence to estimate the terms $\eta_k^{(t+i)}$. If we assume a particular form of the sequence, $\eta_k^{(t)} = \eta_k^{(0)} t^{-\alpha_k}$, then this can be done readily once we have an estimate of $\eta_k^{(0)}$ and $\alpha_k$. However, if each term of the sequence is estimated separately, we need to use model to predict the next terms. We propose a few simple methods in Section~\ref{sec:results}, and evaluate their generalization performance.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{images/user_interface}
  \caption{Web interface for the game.}
  \label{fig:game_interface}
\end{figure*}

@@

\section{Results}
\label{sec:results}

@@

At the beginning of the game, there is a clear exploration phase where players tend make aggressive assignments with their flow distribution. Towards the end of the game, the players consolidate their position and tend to make more conservative flow distribution assignments, as seen in Figure~\ref{fig:player_behavior} as an example.

We can describe the convergence of the system more concretely by evaluating the Rosenthal potential of the game at each turn from Proposition~\ref{prop:potential}. From Figure~\ref{fig:global_potential}, the potential of the system oscillates rapdily in the exploration phase and then converges to the minimal potential. The system also shows robustness seen in Figure~\ref{fig:global_potential} where it recovers from perturbations in the convergence phase.

@@

We can estimate the learning rate of each player with the update in Eq~\ref{eq:estimation_eta}. We obtained these sequence of learning rates for each player and plotted the moving average of length 3 for better visualization in Figure~\ref{fig:moving_average_learning_rate}.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{images/moving_average_learning_rate}
  \caption{Moving average learning rate.}
  \label{fig:moving_average_learning_rate}
\end{figure}

With these estimated sequences of learning rates, we can predict $x_{t+i+1}$ by using the prediction and propagation in \eqref{eq:estimation_step} and \eqref{eq:propagation}. To obtain $\eta^{(t+i)}_k$, we propose three methods. First, we assume $\eta^{(t+i)}_k = \eta^{(t+i-1)}_k$, where $\eta^{t+i-1}_k$ is the estimated learning rate in the previous turn as a baseline method for evaluation. Second, we assume $\eta^{(t+i)}_k$ to be the moving average of the previous estimated etas in the sequence. Finally, we perform linear regression on the estimated $(\eta_k^{(1)}, \dots, \eta_k^{(t-1)})$ to calculate $\eta^{(t+i)}_k$.

We also compare these methods of computing $\eta^{(t+1)}$ with a parameterized learning rate proposal in Section~\ref{sec:estimation}. To obtain the parameterized sequence of $(\eta_k^{(1)}, \eta_k^{(2)}, \dots)$, we perform solve the non-convex optimization problem in \eqref{eq:estimation_alpha_eta0} for each player $k$. Then we use the optimal $\alpha_k, \eta^{(0)}_{k}$ to obtain the learning rate $\eta^{(t+i)}_k$ for our prediction and propagation. These sequences of learning rates can be seen in Figure~\ref{fig:sequence_rate}.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{images/sequence_rate}
  \caption{Estimated parameterized learning rate.}
  \label{fig:sequence_rate}
\end{figure}

We evaluate each method by computing the average Bregman divergence per player per turn. From the results, the farther ahead we look in our prediction, the average Bregman divergence tend to increase for all methods. As the interval of prediction increases, the modeling errors propagate and the quality of our predictions degrade, which is reflected in Figure~\ref{fig:predictions_divergence}.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{images/predictions_divergence}
  \caption{Average Bregman divergence per player and per iteration.}
  \label{fig:predictions_divergence}
\end{figure}

We show an example of the two-step ahead flow distribution predictions made by the parameterized method in Figure~\ref{fig:parameterized} for all of the paths of one player. The flow predictions for each path are generally close to the actual flow distribution from the player. In cases where the player makes an extreme move, the flow predictions tend to underestimate the magnitude of the change in the flow distribution.
